---
title: "Outbreak detection with sntmethods (epi_* verbs)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Outbreak detection with sntmethods (epi_* verbs)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = FALSE
)
```

# Overview

`sntmethods` exposes a four-verb framework for univariate outbreak
detection:

* **`epi_detect()`** — fit one or more detectors against a baseline /
  target split.
* **`epi_evaluate()`** — grade detectors against labelled, simulated, or
  unlabelled ground truth.
* **`epi_ensemble()`** — combine detectors via majority, weighted, or
  score-average voting.
* **`epi_recommend()`** — profile a series and rank candidate detectors.

All four verbs share a single registry of built-in detectors, accessed via
`available_detectors()`. Users can extend the registry via
`register_detector()`.

# A worked example

We simulate a single province's weekly malaria count series over four
years, then inject a synthetic outbreak in the final year. The first three
years are the baseline; the fourth year is the target window we want to
score.

```{r simulate}
library(sntmethods)
library(dplyr)
library(ggplot2)

set.seed(42)

dates <- seq.Date(as.Date("2021-01-04"),
                  as.Date("2024-12-30"),
                  by = "week")
n_w   <- length(dates)
doy   <- as.numeric(format(dates, "%j"))
seasonal <- 100 + 60 * sin(2 * pi * doy / 365.25)
counts   <- rpois(n_w, lambda = seasonal)

# inject a 6-week spike in mid-2024
spike_idx <- which(dates >= as.Date("2024-06-01") &
                   dates <= as.Date("2024-07-15"))
counts[spike_idx] <- counts[spike_idx] +
  rpois(length(spike_idx), lambda = 80)

# truth labels for evaluation
is_outbreak <- rep(FALSE, n_w)
is_outbreak[spike_idx] <- TRUE

df <- tibble::tibble(
  date  = dates,
  cases = counts,
  truth = is_outbreak
)

ggplot(df, aes(date, cases, colour = truth)) +
  geom_line(colour = "grey60") +
  geom_point() +
  theme_minimal()
```

## Step 1: detect

```{r detect}
target_range  <- c(as.Date("2024-01-01"), as.Date("2024-12-30"))
baseline_range <- c(as.Date("2021-01-04"), as.Date("2023-12-25"))

run <- epi_detect(
  data           = df,
  date_col       = "date",
  count_col      = "cases",
  methods        = c("threshold", "ears_c1", "ears_c2",
                     "farrington", "trending"),
  target_range   = target_range,
  baseline_range = baseline_range
)

print(run)
autoplot(run)
```

## Step 2: evaluate

```{r evaluate}
eval <- epi_evaluate(
  detection_run = run,
  truth_col     = "truth",
  loo_method    = TRUE,
  k_sweep       = TRUE,
  cost          = list(c_fp = 1, c_fn = 5)
)

summary(eval)
autoplot(eval)
```

## Step 3: ensemble

```{r ensemble}
ens <- epi_ensemble(
  detection_run = run,
  strategy      = "weighted_vote",
  weights       = eval,
  na_handling   = "as_false"
)

print(ens)
autoplot(ens)
```

## Step 4: recommend

```{r recommend}
recs <- epi_recommend(
  data       = df,
  date_col   = "date",
  count_col  = "cases",
  evaluation = eval,
  tie_break  = "evidence"
)

print(recs)
autoplot(recs)
```

# Extending the framework

Third-party detectors plug in via `register_detector()`. The detector only
needs to honour the four-column predict contract (`date`, `alarm`,
`score`, `upper_threshold`); failure isolation and `method` / `group_id`
columns are added by the framework.

```{r extend}
new_my_zscore <- function(k = 3) {
  sntmethods:::new_epi_detector(method = "my_zscore",
                                params = list(k = k))
}

fit_detector.epi_detector_my_zscore <- function(detector, baseline_data, ...) {
  detector$mu <- mean(baseline_data$cases, na.rm = TRUE)
  detector$sd <- sd(baseline_data$cases, na.rm = TRUE)
  detector$fitted <- TRUE
  detector
}

predict_detector.epi_detector_my_zscore <- function(detector, target_data, ...) {
  z <- (target_data$cases - detector$mu) / detector$sd
  tibble::tibble(
    date            = target_data$date,
    alarm           = z > detector$params$k,
    score           = as.numeric(z),
    upper_threshold = detector$mu + detector$params$k * detector$sd
  )
}

register_detector(
  name              = "my_zscore",
  constructor       = new_my_zscore,
  fit_method        = fit_detector.epi_detector_my_zscore,
  predict_method    = predict_detector.epi_detector_my_zscore
)

available_detectors()
```
